\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2012-03-30 pkg/evtree/Submission_JSS/Revisions/response_revision_14032012.tex

\documentclass[DIN, pagenumber=false, parskip=half,% 
               fromalign=left, fromphone=true,%  
               fromemail=true, fromurl=false, %
               fromlogo=true, fromrule=false, fromrule=afteraddress]{scrlttr2}

\usepackage[latin1]{inputenc}

\usepackage{graphicx}
\usepackage{amsmath}

\setkomavar{fromname}{Thomas Grubinger\\ Department of Medical Statistics, Informatics and Health Economics \\ Innsbruck Medical University}
\setkomavar{fromaddress}{Sch\"opfstrasse 41/1, 6020 Innsbruck, Austria}
\setkomavar{backaddress}{Dept. of Med. Statistics \\ Innsbruck Medical University\\ Sch\"opfstr. 41/1, A-6020 Innsbruck}
\setkomavar{fromphone}{+43 512 9003 70906}
\setkomavar{fromemail}{thomas.grubinger@i-med.ac.at}
\setkomavar{signature}{Thomas Grubinger, Achim Zeileis und Karl-Peter Pfeiffer}
\setkomavar{fromlogo}{ \includegraphics*[width=2.8cm]{logo}}
\setkomavar{subject}[]{Author's response to the review of ``evtree: 
Evolutionary Learning of Globally Optimal Classification and Regression Trees in R'' (JSS 817)}

@@

\textit{The paper did a good job of introducing a new tree-based algorithm overall. A lot of details pertaining to traditional tree models and the concept of evolutionary algorithms were covered, some of which could be clarified further. A clean-up in notation and wording would be greatly beneficial. The code appears easy-to-read and uses parameters consistent with other tree packages.}

We cleaned-up the notation and wording according to the reviewer's recommendations.
\\ \\

\textit{In the end, the authors gave honest comparisons of their methods to the traditional rpart and ctree methods, but it is unclear as to when and why evtree outperforms its competitors apart from the specific chessboard example given in Section 5.2.}

We added the following paragraph to the ``5.1 Benchmark and real-world problemss'', as an explanation as to when and why evtree outperforms its competitors:

### 2012-04-03 pkg/evtree/inst/doc/evtree.Rnw
continued revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@287 255f4249-2263-4417-b1ea-8498779e27a7
small improvements in notation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@288 255f4249-2263-4417-b1ea-8498779e27a7

In the following, a binary tree model with $M$ terminal nodes (which consequently has
$M-1$ internal splits) is denoted by
%
\begin{equation}
  \theta ~=~ (v_{1}, s_{1},...,v_{M-1},s_{M-1} ), \label{eq:tree}
\end{equation}
%
where $r \in \{1,...,M-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
\emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
(for ordered and numeric variables) or a non-empty subset of $\{1, \dots, c\}$
(for a categorical variable with $c$~levels), determining which observations are
sent to the first or second subsample. In the former case, there are $u-1$ possible
split rules if $X_{v_r}$ takes $u$ distinct values; and in the latter case, there
are $2^{c-1} - 1$ possible splits. Thus, the product of all of these combinations
forms all potential elements $\theta$ from $\Theta_M$, the space of conceivable
trees with $M$ terminal nodes. The overall parameter space is then 
$\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
by exluding elements $\theta$ resulting in too small subsamples etc.).

@@

First, the data is loaded and the forward-search trees are grown
with and without depth restriction, visualizing the unrestricted trees
in Figure~\ref{fig:example_rpart_ctree}.
%
<<BBBClub-rpart-ctree, echo=TRUE, eval=FALSE>>=
data("BBBClub", package = "evtree")
library("rpart")
rp  <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
rp2 <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10,
  maxdepth = 2))
ct  <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99)
ct2 <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99,
  maxdepth = 2)
plot(rp)
plot(ct)
@
%
With the objective of building a smaller, but at still accurate tree,
\code{evtree} is constrained to a maximum tree depth of $2$,
see Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree, echo=TRUE, eval=FALSE>>=
set.seed(1090)
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
@
%
<<BBBClub-cache, echo=FALSE, eval=TRUE>>=
if(cache & file.exists("BBBClub-trees.rda")) {
load("BBBClub-trees.rda")
} else {
<<BBBClub-rpart-ctree>>
<<BBBClub-evtree>>
if(cache) {
  save(rp, rp2, ct, ct2, ev, file = "BBBClub-trees.rda")
} else {
  if(file.exists("BBBClub-trees.rda")) file.remove("BBBClub-trees.rda")
}
}
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[p!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth}
<<BBBClub-rpart-plot, fig=TRUE, echo=FALSE, width=8.1, height=7>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
\setkeys{Gin}{width=1\textwidth}
<<BBBClub-ctree-plot, fig=TRUE, echo=FALSE,  width=11.4, height=5.9>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees for customer targeting constructed by \code{rpart}
  (upper panel) and \code{ctree} (lower panel). The target variable is the customer's choice of buying the book.
  The variables used for splitting are the number of art books purchased previously (\code{art}), the number
  of months since the first purchase (\code{first}), the frequency of previous purchases at the Bookbinder's Book Club (\code{freq}), and the customer's \code{gender}.}
\end{figure}
%
The resulting evolutionary tree is printed below and visualized in
Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree-display>>=
plot(ev)
ev
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth} 
<<BBBClub-evtree-plot, fig=TRUE, echo=FALSE, width=7.5, height=4.9>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree for customer targeting constructed by \code{evtree}
   The target variable is the customer's choice of buying the book. The variables used for splitting are
   the number of art books purchased previously (\code{art}), and the number of months since the first purchase
    (\code{first}).}
\end{figure}

>>>>

In the following, a binary tree model with $M$ terminal nodes (which consequently has
$M-1$ internal splits) is denoted by
%
\begin{equation}
  \theta ~=~ (v_{1}, s_{1},...,v_{M-1},s_{M-1} ), \label{eq:tree}
\end{equation}
%
where $v_r \in \{1, \dots ,P\}$ are the splitting \emph{variables} and $s_r$ the
associated \emph{split} rules for the internal \emph{nodes} $r \in \{1,...,M-1\}$.
Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
(for ordered and numeric variables) or a non-empty subset of $\{1, \dots, c\}$
(for a categorical variable with $c$~levels), determining which observations are
sent to the first or second subsample. In the former case, there are $u-1$ possible
split rules if $X_{v_r}$ takes $u$ distinct values; and in the latter case, there
are $2^{c-1} - 1$ possible splits. Thus, the product of all of these combinations
forms all potential elements $\theta$ from $\Theta_M$, the space of conceivable
trees with $M$ terminal nodes. The overall parameter space is then 
$\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
by exluding elements $\theta$ resulting in too small subsamples etc.).

@@

First, the data is loaded and the forward-search trees are grown
with and without depth restriction, visualizing the unrestricted trees
in Figure~\ref{fig:example_rpart_ctree}.
%
<<BBBClub-rpart-ctree, echo=TRUE, eval=FALSE>>=
data("BBBClub", package = "evtree")
library("rpart")
rp  <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
rp2 <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10,
  maxdepth = 2))
ct  <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99)
ct2 <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99,
  maxdepth = 2)
plot(rp)
plot(ct)
@
%
With the objective of building a smaller, but at still accurate tree,
\code{evtree} is constrained to a maximum tree depth of $2$,
see Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree, echo=TRUE, eval=FALSE>>=
set.seed(1090)
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
@
%
<<BBBClub-cache, echo=FALSE, eval=TRUE>>=
if(cache & file.exists("BBBClub-trees.rda")) {
load("BBBClub-trees.rda")
} else {
<<BBBClub-rpart-ctree>>
<<BBBClub-evtree>>
if(cache) {
  save(rp, rp2, ct, ct2, ev, file = "BBBClub-trees.rda")
} else {
  if(file.exists("BBBClub-trees.rda")) file.remove("BBBClub-trees.rda")
}
}
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[p!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth}
<<BBBClub-rpart-plot, fig=TRUE, echo=FALSE, width=8.1, height=7>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
\setkeys{Gin}{width=1\textwidth}
<<BBBClub-ctree-plot, fig=TRUE, echo=FALSE,  width=11.4, height=5.9>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees for customer targeting constructed by \code{rpart}
  (upper panel) and \code{ctree} (lower panel). The target variable is the customer's choice of buying the book.
  The variables used for splitting are the number of art books purchased previously (\code{art}), the number
  of months since the first purchase (\code{first}), the frequency of previous purchases at the Bookbinder's Book Club (\code{freq}), and the customer's \code{gender}.}
\end{figure}
%
The resulting evolutionary tree is printed below and visualized in
Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree-display>>=
plot(ev)
ev
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth} 
<<BBBClub-evtree-plot, fig=TRUE, echo=FALSE, width=7.5, height=4.9>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree for customer targeting constructed by \code{evtree}.
   The target variable is the customer's choice of buying the book. The variables used for splitting are
   the number of art books purchased previously (\code{art}), and the number of months since the first purchase
    (\code{first}).}
\end{figure}
